Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-10-27 analysis/paper_draft2.Rmd
revised

knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, message = FALSE)

@@

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases. (_Flowers v. Mississippi_ -@flowers2019, p. 2243; Wash. Rule 37(g)(v) -@WashR37; Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)). Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by race or gender, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter $b$, to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for $b$. In so doing, we use the _power prior_ [@Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter $b$.  Finally, we use that posterior distribution as the prior for $b$ in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

Following @Kadane2018 and @Barrett2007, we model a "struck-jury" peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under the "struck jury" procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories, so that the number of potential jurors who remain is the sum of number of seats on the jury (plus alternates, if any) and the number of peremptory challenges allotted to both sides. Then, the parties exercise their strikes on those potential jurors, either submitting strikes simultaneously or in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are assigned to seats on the jury.

@@

In some courts, both parties simultaneously exercise their peremptory challenges on the prospective jurors in a venire. This simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process above (see Equation \eqref{eq:model3}), i.e, as equivalent to one party engaging in an uninterrupted sequence of strikes against a subset of prospective jurors eligible to be struck. The premise: Regardless of the order in which a party announced those strikes, the posterior for the bias parameter would be the same.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app using `r R.version$version.string` and shiny [@R; @shiny].

To show how this app might be used by lawyers in real cases, we loaded this app with real strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. To collect this historical strike data, one of us filed a request with the U.S. District Court of Connecticut based on 28 U.S.C. $\S$ 1868.  Thereafter, we received copies (pdf files) of certain jury-selection records associated with [INSERT] trials in that court during this period. These included strike sheets that indicated the ID number of prospective jurors who were struck by peremptory challenge, the order in which they were struck, and which side (prosecutor or defense) struck which juror. Such records also included a tally of answers to juror questionnaires that asked each prospective juror to report their race and gender.

@@

```{r inpts, echo=FALSE}
df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
knitr::kable(df, caption ="Hypothetical Strike Tally")

@@

Finally, for the approach here, incomplete historical strike data presents an issue for inferring bias if such data is missing not at random. In practice, jury-selection records in past cases are hard to get because of court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record. We hope these data challenges decrease over time as judges, attorneys, and jury clerks realize that more historical strike data makes it easier, with the approach described here, to credibly infer or disprove bias under _Batson_ and similar law.

@@

Based on categories in https://casrai.org/credit/.

>>>>

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

@@

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases. (_Flowers v. Mississippi_ -@flowers2019, p. 2243; Wash. Rule 37(g)(v) -@WashR37; Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)). Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the _power prior_ [@Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

Following @Kadane2018 and @Barrett2007, we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury until those seats are filled.

@@

In some courts, both parties simultaneously exercise their peremptory challenges on the prospective jurors subject to strikes. This simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process above (see Equation \eqref{eq:model3}), i.e, as equivalent to one party engaging in an uninterrupted sequence of strikes against a subset of prospective jurors eligible to be struck. The premise: Regardless of the order in which a party announced those strikes, the posterior for the bias parameter would be the same.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and shiny [@R; @shiny].

To show how this app might be used by lawyers in real cases, we loaded this app with real strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. To collect this historical strike data, one of us filed a request with the U.S. District Court of Connecticut based on 28 U.S.C. $\S$ 1868.  Thereafter, we received copies of certain jury-selection records associated with twenty-nine criminal cases in that court during this period. These included strike sheets that indicated the identification number of prospective jurors who were struck by peremptory challenge, the order in which they were struck, and which side (prosecutor or defense) struck which juror. Such records also included a tally of answers to juror questionnaires that asked each prospective juror to report their race and gender.

@@

```{r inpts}
df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
knitr::kable(df, caption ="Hypothetical Strike Tally")

@@

Finally, in practice, historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here for detecting bias if the missing strike data is missing not at random.

@@
